---
title: "04_image_creation"
author: "Aidan Coyle"
date: "5/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is used to create tables and images for display in the final paper and presentation

# Load libraries and data, change data classes

```{r}
library(tidyverse)
library(viridis)

# Load in our data file created in the previous script
crabdat <- read.csv("../output/crab_data_tables/allyears_cleaned.csv")

# Convery columns to factors
crabdat$LOCATION_CODE <- factor(crabdat$LOCATION_CODE)
crabdat$SEX_CODE <- ordered(crabdat$SEX_CODE,
                            levels = c("M", "F"))
crabdat$SHELL_CONDITION_CODE <- ordered(crabdat$SHELL_CONDITION_CODE)
crabdat$BLACKMAT_CODE <- ordered(crabdat$BLACKMAT_CODE)
crabdat$PARASITE_CODE <- ordered(crabdat$PARASITE_CODE)
crabdat$LEG_CONDITION_CODE <- ordered(crabdat$LEG_CONDITION_CODE)

```


### Final Presentation Graphs


```{r graph_creation}
# Shell condition and Hematodinium
ggplot(crabdat, aes(fill = PARASITE_CODE, x = SHELL_CONDITION_CODE)) +
  geom_bar(position = "fill") +
  labs(x = "Months since last molt",
       y = "Proportion") +
  scale_x_discrete(labels = c("0-2", "2-12", "12-24", "24+")) +
  scale_fill_manual(values = viridis(2),
                    name = "Hematodinium",
                    labels = c("Uninfected", "Infected"))



# Depth and Hematodinium
ggplot(crabdat, aes(x = PARASITE_CODE, y = DEPTH_FATHOMS)) +
  geom_violin() +
  scale_y_reverse() +
  labs(x = "Hematodinium", 
       y = "Depth (fathoms)") +
  scale_x_discrete(labels = c("Uninfected", "Infected")) +
  theme_minimal()
  
# Carapace Width  and Hematodinium
ggplot(crabdat, aes(x = PARASITE_CODE, y = WIDTH_MILLIMETERS)) +
  geom_violin() +
  labs(x = "Hematodinium", 
       y = "Carapace width (mm)") +
  scale_x_discrete(labels = c("Uninfected", "Infected")) +
  theme_minimal()
  

  


# Black Mat and Hematodinium
ggplot(crabdat, aes(fill = PARASITE_CODE, x = BLACKMAT_CODE)) +
  geom_bar(position = "fill") +
  labs(x = "Black Mat",
       y = "Proportion") +
  scale_x_discrete(labels = c("Uninfected", "Infected")) +
  scale_fill_manual(values = viridis(2),
                    name = "Hematodinium",
                    labels = c("Uninfected", "Infected"))
  


nrow(crabdat[crabdat$PARASITE_CODE==1,])

```



### Table of Variables and Reasoning


| Variable               | Categorical or Continuous | Reason for choosing                                     | AIC compared to null model |
|------------------------|---------------------------|---------------------------------------------------------|----------------------------|
| Hematodinium infection | Categorical               | Response variable                                       | NA                         |
| Year                   | Continuous                | Random effect. Prevalence may differ by year            | NA                         |
| Location               | Categorical               | Random effect. Prevalence may differ by location        | NA                         |
| Shell condition        | Categorical               | Molting may make crabs vulnerable due to weakened shell | -791.8                     |
| Carapace width         | Continuous                | Life stage may impact vulnerability                     | -104.2                     |
| Black Mat Disease      | Categorical               | May weaken immune system or compete with Hematodinium   | -73.8                      |
| Sex                    | Categorical               | Egg production may impact vulnerability                 | -19.4                      |
| Missing legs?          | Categorical               | Wounds may make crabs vulnerable due to infection       | -8.9                       |
| Julian day             | Continuous                | Prevalence may differ throughout year                   | -3.3                       |



### Table of Avg. Model Output

|                        | Estimate  | Std. Error | Adjusted SE | z value | Pr (>\|z\|) |
|------------------------|-----------|------------|-------------|---------|-------------|
| (Intercept)            | -10.65688 | 11.85864   | 11.85963    | 0.899   | 0.369       |
| BLACKMAT_CODE.L        | -10.12176 | 16.77164   | 16.77304    | 0.603   | 0.546       |
| DEPTH_SCALED           | -0.50887  | 0.05463    | 0.05464     | 9.313   | <2e-16 ***  |
| SHELL_CONDITION_CODE.L | -1.54969  | 0.26261    | 0.26263     | 5.901   | <2e-16 ***  |
| SHELL_CONDITION_CODE.Q | -1.68320  | 0.19912    | 0.19914     | 8.453   | <2e-16 ***  |
| SHELL_CONDITION_CODE.C | 0.64983   | 0.10591    | 0.10592     | 6.135   | <2e-16 ***  |
| WIDTH_SCALED           | 0.20714   | 0.03107    | 0.03108     | 6.665   | <2e-16 ***  |
| DAY_SCALED             | -0.00975  | 0.05200    | 0.05200     | 0.187   | 0.851       |

