---
title: "04_image_creation"
author: "Aidan Coyle"
date: "5/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is used to create tables and images for display in the final paper



### Table of Variables and Reasoning


| Variable               | Categorical or Continuous | Reason for choosing                                                              |
|------------------------|---------------------------|----------------------------------------------------------------------------------|
| Hematodinium infection | Categorical               | Response variable                                                                |
| Year                   | Continuous                | Prevalence may differ by year                                                    |
| Location               | Categorical               | Prevalence may differ by location                                                |
| Carapace width         | Continuous                | Life stage may impact vulnerability                                              |
| Julian day             | Continuous                | Prevalence may differ throughout year                                            |
| Sex                    | Categorical               | Egg production may impact vulnerability                                          |
| Shell condition        | Categorical               | Molting may make crabs vulnerable due to weakened shell                          |
| Black Mat Disease      | Categorical               | May result in decreased immune system or increased resistance due to competition |
| Missing legs?          | Categorical               | Wounds may make crabs vulnerable due to infection                                |


### Table of Avg. Model Output

|                        | Estimate  | Std. Error | Adjusted SE | z value | Pr (>\|z\|) |
|------------------------|-----------|------------|-------------|---------|-------------|
| (Intercept)            | -10.65688 | 11.85864   | 11.85963    | 0.899   | 0.369       |
| BLACKMAT_CODE.L        | -10.12176 | 16.77164   | 16.77304    | 0.603   | 0.546       |
| DEPTH_SCALED           | -0.50887  | 0.05463    | 0.05464     | 9.313   | <2e-16 ***  |
| SHELL_CONDITION_CODE.L | -1.54969  | 0.26261    | 0.26263     | 5.901   | <2e-16 ***  |
| SHELL_CONDITION_CODE.Q | -1.68320  | 0.19912    | 0.19914     | 8.453   | <2e-16 ***  |
| SHELL_CONDITION_CODE.C | 0.64983   | 0.10591    | 0.10592     | 6.135   | <2e-16 ***  |
| WIDTH_SCALED           | 0.20714   | 0.03107    | 0.03108     | 6.665   | <2e-16 ***  |
| DAY_SCALED             | -0.00975  | 0.05200    | 0.05200     | 0.187   | 0.851       |



```{r}

```

